Browsernative Microservices

### 2017-01-25 paper.tex
pdf/tex build

\section{Introduction}\label{introduction}

@@

Introduction: from simplicity to Microservices to extensible web
manifesto to w3c specifications \ldots{} and beyond

Der Erfinder der Programmiersprache Clojure, Rich Hickey, ist ohne
Zweifel eine Koryphäe auf seinem Gebiet, der Strukturierung von
komplexen Systemen. In einer vielbeachteten Keynote aus dem Jahr 2012
geht er auf etymologisch-philosophische Spurensuche nach dem Wort
\textbf{simplicity } aus Sicht eines Softwareentwicklers.\footnote{\href{https://www.youtube.com/watch?v=rI8tNMsozo0\&t=48s}{Rails
  Conf 2012 Keynote: Simplicity Matters by Rich Hickey}} Das Adjektiv
\emph{simple} hat demnach seinen Ursprung im lateinischen Wort
\emph{simplex}, was soviel wie \emph{einfach} oder \emph{einzeln}
bedeutet. In Gegensatz dazu stehen Eigenschaften wie \emph{complex} oder
\emph{multiplex}. Qualitative Software ist, so Hickey, vor allem simpel
- im Prozess, im Design, in der Struktur und in der Entwicklung.

Bezeichnend für diese Keynote ist, dass sie von Rich Hickey im Rahmen
einer Webentwicklerkonferenz gehalten wurde. Software für den Browser
war (und ist) seit langer Zeit maßgeblich geprägt von
\textbf{Komplexität} auf mehreren Ebenen. Im Laufe dieser Bachelorarbeit
werden diese Stru

\begin{itemize}
\tightlist
\item
  Das \emph{Document Object Model} als Rückgrad jeder Webapplikation ist
  hierarchisch strukturiert und deren Elemente damit keinesfalls
  unabhängig in ihrer Darstellung und Reihenfolge.
\item
  JavaScript als single-threaded Skriptsprache lässt sich schlecht ihn
  ihrem Verhalten isolieren, ist Fehleranfällig und hatte lange Zeit nur
  sehr wenig idiomatische Lösungsansätze für komplexe Probleme, wie
  beispielsweise Asynchronität
\item
  CSS ist geprägt von stetigem Überschreiben vorher definierter Regeln
  und verletzt damit Simplizität in ihrer Struktur, Design und dem
  Entwicklungsprozesses auf bester Art und Weise.
\end{itemize}

@@

Nach Abramov existieren zum einen Komponenten, die alleine für das
\textbf{Darstellen von Information} zuständig sind. Diese Komponenten
nehmen keinerlei Einfluss auf Informationen oder anderen Komponenten um
sich herum. Sie sind im wahrsten Sinne passiv und fremdgesteuert über
klar definierte Schnittstellen. Diese Komponenten sind häufig flexibel
einsetzbar und hochgradig wiederwerwendbar. Abramov nennt diese
Komponenten ``Presentational Components''.{[}1{]} Verortet man diese Art
von Komponenten innerhalb des \emph{MVC Pattern}, sind die Komponenten
reine \emph{View-Elemente}.

Im Gegensatz dazu stehen Komponenten, die für die \textbf{Verarbeitung
von Informationen} zuständig sind. Diese so genannten ``Container
Components'' haben oft einen internen Zustand, den sie verändern können
und an ihre Kindkomponenten weiterreichen können.{[}1{]} Im \emph{MVC
Pattern} handelt es sich um die \emph{Controller}. In der funktionalen
Programmiersprache Elm werden diese Elemente \emph{Updater} genannt, was
ihre Funktion noch besser umschreibt.

@@

\section{Microservices}\label{microservices}

@@

In a nutshell a microservice is a small, autonomous service that works
together with other services seamlessly.{[}2, p. 2{]} or with the words
of Fowler and Lewis: ``It is an approach to developing a single
application as a suite of small services, each running in its own
process and communicating with lightweight mechanisms,\ldots{}''{[}3{]}
Microservices incorporate many ideas, like \emph{domain-driven design}
where we try to represent the real world in our code.{[}2, p. 2{]} Or
making use of \emph{continuous delivery} for pushing software rapidly
through \emph{automated deployment} mechanisms in production.{[}3{]}
And, last but not least, microservices utilizes the idea of small teams
with a lot of product knowledge working mostly autonomous on their very
own service with their very own set of tools and techniques.

@@

Talking about microservices in a browsernative context isn't that far
fetched as microservices themselves incorporate many ideas from the web.
Exemplary, microservices often communicate via an HTTP request-response
with resource API's and lightweight messaging.{[}3{]} Nevertheless,
while (server-side) microservices offer wide ranges of technical
possibilities, we must take into account that (client-side) browsers
come with certain constraints and limitations.

Fowler and Lewis issue a call for using services as components. A
component~is regarded as a unit of software that is independently
replaceable and upgradeable. The main advantage of a component in
contrast of library is the possibility of an independent deployment. It
aligns perfectly with the main goal of a microservice architecture to
strip away most of the dependencies in favour of clean
interfaces.{[}3{]}

@@

First of all, from a technical perspective, a microservice reinforces
the \emph{Single Responsibility Principle} defined by Robert C. Martin:
``Gather together those things that change for the same reason and
separate those things that change for different reasons.''{[}4{]} An a
way this principle tackles another often cited design principle of the
\emph{seperation of concerns}. Web Components incorporate this principle
in multiple ways while still remaining flexible.

@@

The concept of microservices incooperates not only a technical
perspective. Microservice patterns are a product of real-world
usage.{[}2, p. 1{]} In a real world we typical have to deal with the so
called \emph{Conway's Law}:

\begin{quote}
``organizations which design systems \ldots{} are constrained to produce
designs which are copies of the communication~structures~of these
organizations''. {[}5{]}
\end{quote}

Following this logic any company, whether it is web-related or not,
should be devided in units grouped around a destinct business service to
optimise the workflow. Fowler and Lewis outlines this approach as an
``alignment of business capabilities''{[}3{]} While this kind of
structure may be true for companies like Google or Amazon, there is a
vast majority of companies developing for the web which are grouped
around tasks.{[}6{]} A very common structure is formed by the technology
stack (UX Designers, Frontend- \& Backend Developers) or by separating
teams along the product lifecycle (development, testing, deployment).

@@

Depending on the context, some people argue for only two specs which
essentially make it possible to create a scoped component but not caring
too much on it's distribution{[}7{]}. Some people prefer the three specs
{[}8{]}, but the majority advocating the four specs variant, which is
listed on \sout{the official}
\href{http://webcomponents.org}{webcomponents.org} website. For the
purpose of this article, all four specs will be discussed briefly to
provide a rough understanding. It is not meant to cover all bits and
pieces.

@@

Most obvious, this spec relies on the new \emph{ES6 Class Syntax} in
favor of the original prototype-based inheritance model.
``Extending~\texttt{HTMLElement}~ensures the custom element inherits the
entire DOM API and means any properties/methods that you add to the
class become part of the element's DOM interface.''{[}9{]} Like any
other \emph{ES6 class}, \emph{custom element}s can be sub-classed
further on with the typical \texttt{extends} inheritance.

@@

\begin{itemize}
\tightlist
\item
  \texttt{connectedCallback()}\\
  Called upon the time of \emph{connecting or upgrading the node} which
  means the moment the node is inserted and rendered inside the DOM.
  Typically this block of code contains setup code, such as fetching
  resources or rendering.{[}9{]} For performance issues it's highly
  preferable to put much code in here.
\item
  \texttt{disconnectedCallback()}\\
  Called upon the time of \emph{node removal}. Cleanup code like
  removing eventListeners or disconnecting websockets can be put here.
\item
  \texttt{attributeChangedCallback(attrName,\ oldVal,\ newVal)}\\
  This method provides an \emph{Onchange handler} that runs for certain
  attributes called with three values as defined in the signature. It is
  meant to control an elements' transition from on \texttt{oldVal} to a
  \texttt{newVal}. Due to performance issues, this callback is only
  triggered for attributes registered in an \emph{observedAttributes}
  array.
\item
  \texttt{adoptedCallback()}\\
  Called when moving the node \emph{between documents}.
\end{itemize}

@@

\begin{Shaded}
\begin{Highlighting}[]
\OperatorTok{>} \VariableTok{main}\NormalTok{.}\AttributeTok{js}
\KeywordTok{class} \NormalTok{HelloWorld }\KeywordTok{extends} \NormalTok{HTMLElement }\OperatorTok{\{}
  \AttributeTok{constructor}\NormalTok{() }\OperatorTok{\{}\NormalTok{...}\OperatorTok{\}}
  \NormalTok{set }\AttributeTok{sayhello}\NormalTok{(val) }\OperatorTok{\{}
    \KeywordTok{this}\NormalTok{.}\AttributeTok{_hello} \OperatorTok{=} \NormalTok{val}\OperatorTok{;}
    \VariableTok{console}\NormalTok{.}\AttributeTok{log}\NormalTok{(}\KeywordTok{this}\NormalTok{.}\AttributeTok{_hello}\NormalTok{)}\OperatorTok{;}
  \OperatorTok{\}}
  \NormalTok{get }\AttributeTok{sayhello}\NormalTok{() }\OperatorTok{\{}
    \ControlFlowTok{return} \KeywordTok{this}\NormalTok{.}\AttributeTok{_hello}\OperatorTok{;}
  \OperatorTok{\}}
\OperatorTok{\}}
\VariableTok{customElements}\NormalTok{.}\AttributeTok{define}\NormalTok{(}\StringTok{'hello-world'}\OperatorTok{,} \NormalTok{HelloWorld)}\OperatorTok{;}
\KeywordTok{var} \NormalTok{el }\OperatorTok{=} \KeywordTok{new} \AttributeTok{HelloWorld}\NormalTok{()}\OperatorTok{;}
\VariableTok{el}\NormalTok{.}\AttributeTok{sayhello} \OperatorTok{=} \StringTok{"earth"}\OperatorTok{;}
\VariableTok{el}\NormalTok{.}\AttributeTok{sayhello}\OperatorTok{;}\CommentTok{//"earth"}
\end{Highlighting}
\end{Shaded}

While getters and setters work great in the JS world they fail crossing
the boundaries to the corresponding HTML node. Declaring
\texttt{\textless{}hello-world\ sayhello="mars"\textgreater{}\textless{}/hello-world\textgreater{}}
would't work in the previous setup. A common workaround is archived by
using the previous mentioned \texttt{attributeChangedCallback} lifecycle
method to \textbf{reflect changing HTML attributes to JS} and/or map JS
attributes to HTML with \texttt{this.setAttributes(...)} respectively.
On \textbf{insertion time} html attributes might raise their hand with
\texttt{this.hasAttributes(...)} and \texttt{this.getAttributes(...)}.
Native DOM properties will reflect their values between HTML and JS
automatically.{[}10, Para. 2.6.1{]}

@@

One aspect didn't mentioned yet is the possibility of creating
sub-classes of \textbf{build-in elements} by extending the native
Interfaces like the \texttt{HTMLButtonElement} interface. While this
functionality is perfectly spec'd it is strongly rejected by some
browser vendors.\footnote{Discussion on the topic:
  https://github.com/w3c/webcomponents/issues/509} Most likely the spec
will change in future in one or other way on this issue and therefore
\textbf{customized build-in elements} left out of this paper
intentionally.

@@

\begin{Shaded}
\begin{Highlighting}[]
\OperatorTok{>} \VariableTok{main}\NormalTok{.}\AttributeTok{js}
\KeywordTok{class} \NormalTok{HelloWorld }\KeywordTok{extends} \NormalTok{HTMLElement }\OperatorTok{\{}
  \AttributeTok{constructor}\NormalTok{() }\OperatorTok{\{}
    \NormalTok{...}
    \KeywordTok{const} \NormalTok{shadowRoot }\OperatorTok{=}
          \KeywordTok{this}\NormalTok{.}\AttributeTok{attachShadow}\NormalTok{(}\OperatorTok{\{}\DataTypeTok{mode}\OperatorTok{:} \StringTok{'open'}\OperatorTok{\}}\NormalTok{)}\OperatorTok{;}
    \VariableTok{shadowRoot}\NormalTok{.}\AttributeTok{innerHTML} \OperatorTok{=} \StringTok{'<p>hello</p>'}\OperatorTok{;}
  \OperatorTok{\}}
\OperatorTok{\}}
\end{Highlighting}
\end{Shaded}

@@

Using the \emph{functional selector} of \texttt{:host()} or even the
only-functional \texttt{:host-contest()} allows the creation of
\textbf{context-aware custom elements}. A possible use case would be
``theming'' a component (example taken from {[}11{]}):

@@

Wrapping up this section, \emph{shadow DOM} provides a non-hacky way to
create uniform looking \emph{custom elements} and even enhance styling
possibilities without adding much overhead. Still, for smaller
components with only one or two child nodes, just a little styling
and/or no structured redistribution a \emph{shadow DOM} might be to hard
to reason about. Eventually it all depends on the question of ``how hard
is it to implement it without shadow DOM'' - which can't be answered
universally. For a more in-depth guide, Google Engineer Eric Bidelman
wroten a great primer on \emph{shadow DOM}{[}11{]}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Querying the template node in question}\\
  \texttt{const\ t\ =\ document.querySelector(\textquotesingle{}\#helloworld\textquotesingle{});}
\item
  \textbf{Preparing the templates' content}\\
  The templates' \texttt{content} property contains all nodes a
  \emph{DocumentFragment} object. Handling should be straight forward\\
  \texttt{t.content.querySelector(\textquotesingle{}img\textquotesingle{}).src\ =\ \textquotesingle{}logo.png\textquotesingle{};}
\item
  \textbf{Optional: Cloning the \emph{DocumentFragment} for multiple
  use}\\
  \texttt{const\ clone\ =\ t.content.cloneNode("deep");}
\item
  \textbf{Appending the clone/original to destination}\\
  \texttt{document.body.appendChild(clone);}
\end{enumerate}

As easy and minimal \emph{HTML templates} are, they're missing out a
crucial feature other template implementations usually have. As
templates are basically just dump containers for HTML Markup, there is
no way to define some logic as \textbf{placeholders} where content
should appear. Of course, with heavy use of JS things could be modeled
this way. But the idiomatic way tends more towards a \emph{Shadow DOM \&
HTML Templates} symbiosis.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{> hello-component.html}
\KeywordTok{<hello-world>}
  \KeywordTok{<p}\OtherTok{ id=}\StringTok{"sendto"}\OtherTok{ slot=}\StringTok{"placeholder"}\KeywordTok{>}
    \NormalTok{Hello World    }
  \KeywordTok{</p>}
\KeywordTok{</hello-world>}

\KeywordTok{<template}\OtherTok{ id=}\StringTok{"hello"}\KeywordTok{>}
  \KeywordTok{<style>}
    \FloatTok{#stylewrapper} \KeywordTok{\{}
      \KeywordTok{font-weight:} \DataTypeTok{bold}\KeywordTok{;}
      \KeywordTok{color:} \NormalTok{orange}\KeywordTok{;}
    \KeywordTok{\}}
  \KeywordTok{</style>}
  \CommentTok{<!-- CONTENT -->}
  \KeywordTok{<div}\OtherTok{ id=}\StringTok{"stylewrapper"}\KeywordTok{>}
    \KeywordTok{<slot}\OtherTok{ name=}\StringTok{"placeholder"}\KeywordTok{>}
      \NormalTok{Named placeholder}
    \KeywordTok{</slot>}
  \KeywordTok{</div>}
\KeywordTok{</template>}

\KeywordTok{<script>}
  \KeywordTok{class} \NormalTok{HelloWorld }\KeywordTok{extends} \NormalTok{HTMLElement }\OperatorTok{\{}
    \AttributeTok{constructor}\NormalTok{() }\OperatorTok{\{}
      \KeywordTok{super}\NormalTok{()}\OperatorTok{;}
      \KeywordTok{const} \NormalTok{shadowRoot }\OperatorTok{=}
        \KeywordTok{this}\NormalTok{.}\AttributeTok{attachShadow}\NormalTok{(}\OperatorTok{\{}\DataTypeTok{mode}\OperatorTok{:} \StringTok{'open'}\OperatorTok{\}}\NormalTok{)}\OperatorTok{;}
      \KeywordTok{const} \NormalTok{template }\OperatorTok{=}
        \VariableTok{document}\NormalTok{.}\AttributeTok{querySelector}\NormalTok{(}\StringTok{'#hello'}\NormalTok{).}\AttributeTok{content}\OperatorTok{;}

      \KeywordTok{this}\NormalTok{.}\VariableTok{shadowRoot}\NormalTok{.}\AttributeTok{appendChild}\NormalTok{(template)}\OperatorTok{;}
    \OperatorTok{\}}
  \OperatorTok{\}}
  \VariableTok{customElements}\NormalTok{.}\AttributeTok{define}\NormalTok{(}\StringTok{'hello-world'}\OperatorTok{,} \NormalTok{HelloWorld)}\OperatorTok{;}
\OperatorTok{<}\SpecialStringTok{/script>}
\end{Highlighting}
\end{Shaded}

>>>>

\section{Simplicity and the state of the
web}\label{simplicity-and-the-state-of-the-web}

@@

Computers can scale, humans can't. Ever since, a program or complex
system made by humans has been constrained by our mental capabilities.
Like in the analogy of juggling balls, our brain can just ``juggle'' a
few things at a time. Rich Hickey, the inventor of the programming
language \emph{Clojure} gave an inspirational keynote on the topic of
\textbf{simplicity}.\footnote{\href{https://www.youtube.com/watch?v=rI8tNMsozo0\&t=46s}{Rails
  Conf 2012 Keynote: Simplicity Matters by Rich Hickey}} In every sphere
of a humans life, simplicity aligns perception with our mental
capacities. Understanding the word \textbf{simple} alone will outline
the this paper on its own.

Derived from the ancient Latin word \textbf{simplex}, simple can be
understood as ``literally, uncompounded or onefold''\footnote{\href{http://www.etymonline.com/index.php?term=simple}{Etymology
  Dictionary}} which points directly to the unidimensional aspect. While
complexity describes the multilayered und entangled nature of
conditions, simplicity empowers the human brain to reason about issues
in a straightforward manner. It certainly has some overlapping's with
\emph{easy}, but while \emph{easy} is more of a relative nature, simple
can be laid out as a objective manner - a goal software development
should pursuit and praise.

Software development is undoubtedly rich in complexity and full of
subtle pitfalls for the human brain. In a typical scenario, a piece of
software evolves over time in one or another \sout{opinionated}
direction. Layers of new abstractions wrestling with old legacy
abstractions and mutating objects will get out of control. Subtle bugs
start to creep in. Eventually the small piece of software ends up in a
highly complected monolith which will determine future design decisions
to a painful degree. Any company will be highly determined by the
current state trying to ``keep the lights on''.

On the other hand, any truly agile system architecture is laid out in a
fine-grained manner. As Rich Hickey argues, design decisions should be
made under the \textbf{impression of extending, substitution, moving,
combining and repurposing}. The ability to reason about the program at
any given time is crucial for future decisions and implementations.
Recalling again the unidimensional nature of simplicity.

The \emph{state of the web} is certainly a different kind of complex
multidimensional beast. ``The web'' is coined to be meant ``everything
that runs in the browser''. While simplicity in the backend is mostly a
matter of principles, any frontend developer is highly restricted on the
entangled nature of the browser platform. Every call to an webpage
results in a single \textbf{global DOM tree} managed by the browser
engine. In the early days of the web a typical static page contained
just a bunch of text nodes and some styling. Nowadays a typical webapp
contains text nodes, functionality and styling from different resources
and is expected to work highly dynamic and frictionless.

In the last four years the average transfer size of a webpage doubled to
currently around 2.5 MB.{[}\^{}http {]}Subtracting images, fonts or
other content the size of HTML, CSS and JS sums up to a total average of
550 kb. One character weights around 1 byte which means an average
webpage is delivering 550.000 character or around 125 pages of
single-spaced text. Frederic Filloux calculated the ratio of real
content on different newspaper websites and came to the conclusion, that
only round about 5-6 \% of the transferred characters made for
consumption.{[}1{]}

Having an 95 \% overhead is rather undesirable for both the consumers
and creators of the website. Since it's a widespread problem without a
single point of failure one can argue the platform itself is the
failure. In the recent years many frameworks, libraries and
methodologies tried to simplify the way of building user interfaces by
scoping assets and design rules into maintainable components. One of the
most adopted technique was the \textbf{virtual DOM} approach essential
ditching the old rule of separated HTML, CSS and JS in favor of
\textbf{scoped JS components} (containing HTML and CSS). While
frameworks are a valid approach to solve the problem they remain highly
opinionated.

In 2013 thinkers, creators and browser vendors joined together to
propose \emph{The Extensible Web Manifesto}.\footnote{\href{https://extensiblewebmanifesto.org/}{The
  Extensible Web Manifesto}} The claim of the Manifesto is to enhance
the current web with new low-level capacities. Those capacities should
empower creators of the web to write more declarative code and therefore
overcome known bottlenecks and artificial abstractions. Four years
later, the new capabilities of JavaScript leapfrogged and many new
low-level APIs brought to life. With the new APIs at hand a webdeveloper
can create scoped and highly reusable microservices without additional
libraries even directly in the browser console.

\emph{Disclaimer:} Currently, many of the new standards proposed
throughout this paper can only tried and tested inside Google Chrome or
Opera. As the innovation cycle is pretty short it is expected to see
those standards go live in all major browsers soon.In the meantime all
of those functionalities can be safely \textbf{polyfilled}.

\subsection{Microservices}\label{microservices}

Microserve

untangle the web

@@

Nach Abramov existieren zum einen Komponenten, die alleine für das
\textbf{Darstellen von Information} zuständig sind. Diese Komponenten
nehmen keinerlei Einfluss auf Informationen oder anderen Komponenten um
sich herum. Sie sind im wahrsten Sinne passiv und fremdgesteuert über
klar definierte Schnittstellen. Diese Komponenten sind häufig flexibel
einsetzbar und hochgradig wiederwerwendbar. Abramov nennt diese
Komponenten ``Presentational Components''.{[}2{]} Verortet man diese Art
von Komponenten innerhalb des \emph{MVC Pattern}, sind die Komponenten
reine \emph{View-Elemente}.

Im Gegensatz dazu stehen Komponenten, die für die \textbf{Verarbeitung
von Informationen} zuständig sind. Diese so genannten ``Container
Components'' haben oft einen internen Zustand, den sie verändern können
und an ihre Kindkomponenten weiterreichen können.{[}2{]} Im \emph{MVC
Pattern} handelt es sich um die \emph{Controller}. In der funktionalen
Programmiersprache Elm werden diese Elemente \emph{Updater} genannt, was
ihre Funktion noch besser umschreibt.

@@

\section{Microservices}\label{microservices-1}

@@

In a nutshell a microservice is a small, autonomous service that works
together with other services seamlessly.{[}3, p. 2{]} or with the words
of Fowler and Lewis: ``It is an approach to developing a single
application as a suite of small services, each running in its own
process and communicating with lightweight mechanisms,\ldots{}''{[}4{]}
Microservices incorporate many ideas, like \emph{domain-driven design}
where we try to represent the real world in our code.{[}3, p. 2{]} Or
making use of \emph{continuous delivery} for pushing software rapidly
through \emph{automated deployment} mechanisms in production.{[}4{]}
And, last but not least, microservices utilizes the idea of small teams
with a lot of product knowledge working mostly autonomous on their very
own service with their very own set of tools and techniques.

@@

Talking about microservices in a browsernative context isn't that far
fetched as microservices themselves incorporate many ideas from the web.
Exemplary, microservices often communicate via an HTTP request-response
with resource API's and lightweight messaging.{[}4{]} Nevertheless,
while (server-side) microservices offer wide ranges of technical
possibilities, we must take into account that (client-side) browsers
come with certain constraints and limitations.

Fowler and Lewis issue a call for using services as components. A
component~is regarded as a unit of software that is independently
replaceable and upgradeable. The main advantage of a component in
contrast of library is the possibility of an independent deployment. It
aligns perfectly with the main goal of a microservice architecture to
strip away most of the dependencies in favour of clean
interfaces.{[}4{]}

@@

First of all, from a technical perspective, a microservice reinforces
the \emph{Single Responsibility Principle} defined by Robert C. Martin:
``Gather together those things that change for the same reason and
separate those things that change for different reasons.''{[}5{]} An a
way this principle tackles another often cited design principle of the
\emph{seperation of concerns}. Web Components incorporate this principle
in multiple ways while still remaining flexible.

@@

The concept of microservices incooperates not only a technical
perspective. Microservice patterns are a product of real-world
usage.{[}3, p. 1{]} In a real world we typical have to deal with the so
called \emph{Conway's Law}:

\begin{quote}
``organizations which design systems \ldots{} are constrained to produce
designs which are copies of the communication~structures~of these
organizations''. {[}6{]}
\end{quote}

Following this logic any company, whether it is web-related or not,
should be devided in units grouped around a destinct business service to
optimise the workflow. Fowler and Lewis outlines this approach as an
``alignment of business capabilities''{[}4{]} While this kind of
structure may be true for companies like Google or Amazon, there is a
vast majority of companies developing for the web which are grouped
around tasks.{[}7{]} A very common structure is formed by the technology
stack (UX Designers, Frontend- \& Backend Developers) or by separating
teams along the product lifecycle (development, testing, deployment).

@@

Depending on the context, some people argue for only two specs which
essentially make it possible to create a scoped component but not caring
too much on it's distribution{[}8{]}. Some people prefer the three specs
{[}9{]}, but the majority advocating the four specs variant, which is
listed on \sout{the official}
\href{http://webcomponents.org}{webcomponents.org} website. For the
purpose of this article, all four specs will be discussed briefly to
provide a rough understanding. It is not meant to cover all bits and
pieces.

@@

Most obvious, this spec relies on the new \emph{ES6 Class Syntax} in
favor of the original prototype-based inheritance model.
``Extending~\texttt{HTMLElement}~ensures the custom element inherits the
entire DOM API and means any properties/methods that you add to the
class become part of the element's DOM interface.''{[}10{]} Like any
other \emph{ES6 class}, \emph{custom element}s can be sub-classed
further on with the typical \texttt{extends} inheritance.

@@

\begin{itemize}
\tightlist
\item
  \texttt{connectedCallback()}\\
  Called upon the time of \emph{connecting or upgrading the node} which
  means the moment the node is rendered inside the DOM. Typically this
  method is called straight after the \texttt{constructor()} on insert.
  This block of code contains setup code, such as fetching resources or
  rendering elements according to attributes.{[}10{]} For performance
  issues it's highly preferable to put much code in here.
\item
  \texttt{disconnectedCallback()}\\
  Called upon the time of \emph{node removal}. Cleanup code like
  removing eventListeners or disconnecting websockets can be put here.
\item
  \texttt{attributeChangedCallback(attrName,\ oldVal,\ newVal)}\\
  This method provides an \emph{Onchange handler} that runs for certain
  attributes called with three values as defined in the signature. It is
  meant to control an elements' transition from on \texttt{oldVal} to a
  \texttt{newVal}. Due to performance issues, this callback is only
  triggered for attributes registered in an \emph{observedAttributes}
  array.
\item
  \texttt{adoptedCallback()}\\
  Called when moving the node \emph{between documents}.
\end{itemize}

@@

\begin{Shaded}
\begin{Highlighting}[]
\OperatorTok{>} \VariableTok{main}\NormalTok{.}\AttributeTok{js}
\KeywordTok{class} \NormalTok{HelloWorld }\KeywordTok{extends} \NormalTok{HTMLElement }\OperatorTok{\{}
  \AttributeTok{constructor}\NormalTok{() }\OperatorTok{\{}\NormalTok{...}\OperatorTok{\}}
  \NormalTok{set }\AttributeTok{sayhello}\NormalTok{(val) }\OperatorTok{\{}
    \KeywordTok{this}\NormalTok{.}\AttributeTok{_hello} \OperatorTok{=} \NormalTok{val}\OperatorTok{;}
    \VariableTok{console}\NormalTok{.}\AttributeTok{log}\NormalTok{(}\KeywordTok{this}\NormalTok{.}\AttributeTok{_hello}\NormalTok{)}\OperatorTok{;}
  \OperatorTok{\}}
  \NormalTok{get }\AttributeTok{sayhello}\NormalTok{() }\OperatorTok{\{}
    \ControlFlowTok{return} \KeywordTok{this}\NormalTok{.}\AttributeTok{_hello}\OperatorTok{;}
  \OperatorTok{\}}
\OperatorTok{\}}\NormalTok{)}\OperatorTok{;}
\VariableTok{customElements}\NormalTok{.}\AttributeTok{define}\NormalTok{(}\StringTok{'hello-world'}\OperatorTok{,} \NormalTok{HelloWorld)}\OperatorTok{;}
\CommentTok{// Instantiation}
\KeywordTok{var} \NormalTok{el }\OperatorTok{=} \KeywordTok{new} \AttributeTok{HelloWorld}\NormalTok{()}\OperatorTok{;}
\VariableTok{el}\NormalTok{.}\AttributeTok{sayhello} \OperatorTok{=} \StringTok{"earth"}\OperatorTok{;}
\VariableTok{el}\NormalTok{.}\AttributeTok{sayhello}\OperatorTok{;}\CommentTok{//"earth"}
\end{Highlighting}
\end{Shaded}

While getters and setters work great in the JS world they fail crossing
the boundaries to the corresponding HTML node. Declaring
\texttt{\textless{}hello-world\ sayhello="mars"\textgreater{}\textless{}/hello-world\textgreater{}}
would't work in the previous setup. A common workaround is archived by
using the previous mentioned \texttt{attributeChangedCallback} lifecycle
method to \textbf{reflect changing HTML attributes to JS} and/or map JS
attributes to HTML with \texttt{this.setAttributes(...)} respectively.
On \textbf{insertion time} html attributes might raise their hand with
\texttt{this.hasAttributes(...)} and \texttt{this.getAttributes(...)}.
Native DOM properties will reflect their values between HTML and JS
automatically.{[}11, Para. 2.6.1{]}

@@

One aspect didn't mentioned yet is the possibility of creating
sub-classes of \textbf{build-in elements} by extending the native
Interfaces like the \texttt{HTMLButtonElement} interface. While this
functionality is perfectly spec'd it is strongly rejected by some
browser vendors.\footnote{https://github.com/w3c/webcomponents/issues/509}
Most likely the spec will change in future in one or other way on this
issue and therefore \textbf{customized build-in elements} left out of
this paper intentionally.

@@

\begin{Shaded}
\begin{Highlighting}[]
\OperatorTok{>} \VariableTok{main}\NormalTok{.}\AttributeTok{js}
\KeywordTok{class} \NormalTok{HelloWorld }\KeywordTok{extends} \NormalTok{HTMLElement }\OperatorTok{\{}
  \AttributeTok{constructor}\NormalTok{() }\OperatorTok{\{}
    \NormalTok{...}
    \KeywordTok{this}\NormalTok{.}\AttributeTok{attachShadow}\NormalTok{(}\OperatorTok{\{}\DataTypeTok{mode}\OperatorTok{:} \StringTok{'open'}\OperatorTok{\}}\NormalTok{)}\OperatorTok{;}
    \VariableTok{shadowRoot}\NormalTok{.}\AttributeTok{innerHTML} \OperatorTok{=} \StringTok{'<p>hello</p>'}\OperatorTok{;}
  \OperatorTok{\}}
\OperatorTok{\}}
\end{Highlighting}
\end{Shaded}

@@

Using the \emph{functional selector} of \texttt{:host()} or even the
only-functional \texttt{:host-contest()} allows the creation of
\textbf{context-aware custom elements}. A possible use case would be
``theming'' a component (example taken from {[}12{]}):

@@

Wrapping up this section, \emph{shadow DOM} provides a non-hacky way to
create uniform looking \emph{custom elements} and even enhance styling
possibilities without adding much overhead. Still, for smaller
components with only one or two child nodes, just a little styling
and/or no structured redistribution a \emph{shadow DOM} might be to hard
to reason about. Eventually it all depends on the question of ``how hard
is it to implement it without shadow DOM'' - which can't be answered
universally. For a more in-depth guide, Google Engineer Eric Bidelman
wroten a great primer on \emph{shadow DOM}{[}12{]}.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Querying the template node in question}\\
  const node = document.querySelector(`template');
\item
  \textbf{Parsing the content and preparing the templates' content}\\
  const content = node.content;\\
  -\textgreater{} Returns a \emph{DocumentFragment} object. Handling is
  straight forward\\
  content.querySelector(`img').src = `logo.png';
\item
  \textbf{Optional: Cloning the \emph{DocumentFragment} for multiple
  use}\\
  const clone = content.cloneNode(``deep'');
\item
  \textbf{Appending the clone/original to destination}\\
  document.body.appendChild(clone);
\end{enumerate}

As easy and minimal \emph{HTML templates} are, they're missing out a
crucial feature other template implementations usually have. As
templates are basically just dump containers for HTML Markup, there is
no way to define some logic as \textbf{placeholders} where content
should appear. Of course, with heavy use of JS things could be modeled
this way. The idiomatic way tends more towards a \emph{Shadow DOM \&
HTML templates} symbiosis.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{> index.html}
\KeywordTok{<hello-world>}
  \KeywordTok{<p}\OtherTok{ id=}\StringTok{"sendto"}\OtherTok{ slot=}\StringTok{"placeholder"}\KeywordTok{>}
    \NormalTok{Hello World Web Component    }
  \KeywordTok{</p>}
\KeywordTok{</hello-world>}
\CommentTok{<!-- COMPONENT STARTS HERE -->}
\KeywordTok{<template}\OtherTok{ id=}\StringTok{"hello"}\KeywordTok{>}
  \CommentTok{<!-- STYLES -->}
  \KeywordTok{<style>}
    \FloatTok{#stylewrapper} \KeywordTok{\{}
      \KeywordTok{font-weight:} \DataTypeTok{bold}\KeywordTok{;}
      \KeywordTok{color:} \NormalTok{orange}\KeywordTok{;}
    \KeywordTok{\}}
  \KeywordTok{</style>}
  \CommentTok{<!-- CONTENT -->}
  \KeywordTok{<div}\OtherTok{ id=}\StringTok{"stylewrapper"}\KeywordTok{>}
    \KeywordTok{<slot}\OtherTok{ name=}\StringTok{"placeholder"}\KeywordTok{>}
      \NormalTok{Named placeholder}
    \KeywordTok{</slot>}
  \KeywordTok{</div>}
\KeywordTok{</template>}

\KeywordTok{<script>}
  \CommentTok{// Switched to anonymous class notation}
  \CommentTok{// for keeping associated code together.}
  \VariableTok{customElements}\NormalTok{.}\AttributeTok{define}\NormalTok{(}\StringTok{'hello-world'}\OperatorTok{,}
    \KeywordTok{class} \KeywordTok{extends} \NormalTok{HTMLElement }\OperatorTok{\{}
     \AttributeTok{constructor}\NormalTok{() }\OperatorTok{\{}
       \KeywordTok{super}\NormalTok{()}\OperatorTok{;}
       \KeywordTok{this}\NormalTok{.}\AttributeTok{attachShadow}\NormalTok{(}\OperatorTok{\{}\DataTypeTok{mode}\OperatorTok{:} \StringTok{'open'}\OperatorTok{\}}\NormalTok{)}\OperatorTok{;}
       \KeywordTok{const} \NormalTok{template }\OperatorTok{=}
         \VariableTok{document}\NormalTok{.}\AttributeTok{querySelector}\NormalTok{(}\StringTok{'#hello'}\NormalTok{)}\OperatorTok{;}
       \KeywordTok{this}\NormalTok{.}\AttributeTok{shadowRoot}
         \NormalTok{.}\AttributeTok{appendChild}\NormalTok{(}\VariableTok{template}\NormalTok{.}\AttributeTok{content}\NormalTok{)}\OperatorTok{;}
     \OperatorTok{\}}
   \OperatorTok{\}}\NormalTok{)}\OperatorTok{;}
\OperatorTok{<}\SpecialStringTok{/script>}
\end{Highlighting}
\end{Shaded}
